---
title: "Advanced Data Visualisation with R"
subtitle: "Writing `ggplot2` extensions"
author: "Emi Tanaka"
email: "emi.tanaka@monash.edu"
date: "8th Dec 2021 @ Statistical Society of Australia Canberra Branch | Zoom"
length: "360 minutes"
color_theme: "yellow"
titleslide_bgimg: "images/bg4.jpg"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/font-awesome-all.css"
      - "assets/tachyons-addon.css"
      - "assets/animate.css"
      - "assets/fira-code.css"
      - "assets/boxes.css"
      - "assets/styles.css"
      - "assets/monash-brand.css"
      - "assets/monash-fonts.css"
      - "assets/slide-types.css"
      - "assets/table.css"
      - "assets/custom.css"
    self_contained: false 
    seal: false 
    chakra: 'libs/remark-latest.min.js'
    lib_dir: libs
    #mathjax: "assets/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: github
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
---


```{r, include = FALSE}
current_file <- knitr::current_input()
figpath <- paste0("images/", gsub("[.]Rmd", "", current_file), "/")
knitr::opts_chunk$set(fig.align = "center",
                      fig.path = figpath,
                      fig.retina = 2,
                      cache = FALSE,
                      cache.path = "cache/day1-session3/")
library(ggplot2)
theme_set(theme(text = element_text(size = 18)))
```
```{r titleslide, child="assets/titleslide.Rmd"}
```

---

# Dissecting the `ggplot` object 


```{r}
library(ggplot2)
g <- ggplot(cars, aes(speed, dist)) +
  geom_point()  
```

.flex.mt3[
.w-35.pr3[
```{r ggcars, fig.height = 5, fig.width = 5}
g
```

]
.w-60[
```{r str-chickwts, eval = FALSE}
str(g)
```
.scroll-sign.f5[
.overflow-scroll.h5[
```{r str-chickwts, echo = FALSE}
```
]]
{{content}}
]
]

--

* Notice that a layer, `geom`, `position`, `stat`, `scales`, `coordinates` and `facet` are **`ggproto` objects**


---


class: transition 

# `ggproto`

---


# `ggproto` basics

* `ggplot2` makes heavy use of **prototype-based programming** 
--

* `ggproto` is a custom build class system made specically for `ggplot`
--

* The system is similar to `R6Class` in the `R6` package that allow inheritance and method access from parent classes
--



.flex[
.w-50.pr3[
.f5[
```{r}
OzCovidTracker <- ggproto("OzCovidTracker", NULL, 
                cases = 0,
                location = "Australia",
                last_update = NA,
                add = function(self, cases = 0) {
                  self$cases <- self$cases + cases
                  self$last_update <- Sys.Date()
                },
                reset = function(self) {
                  self$cases <- 0
                })
```

]
]
.w-50.f5[
{{content}}
]
]

--

```{r}
OzCovidTracker
```

{{content}}

--

```{r}
OzCovidTracker$add(cases = 219120)
```
{{content}}

--

```{r}
OzCovidTracker$cases
```
{{content}}

--

```{r}
OzCovidTracker$add(cases = 80)
```
{{content}}

--

```{r}
OzCovidTracker$cases
```

---

# `ggproto` inheritance

```{r}
VicCovidTracker <- ggproto("VicCovidTracker", OzCovidTracker, location = "Victoria")
```

--

```{r}
VicCovidTracker$reset()
VicCovidTracker
```

--

```{r}
VicCovidTracker$add(128849)
```

--

```{r}
VicCovidTracker$cases
```

---

# Creating a new `ggproto` object


* You should not be creating a new `ggroto` object from scratch
--

* You should inherit existing `ggproto` objects as outlined below:


New `ggproto` | Parent `ggproto`
--- | ---
`geom` | `ggplot2::Geom`
`position` | `ggplot2::Position`
`stat` | `ggplot2::Stat`
`scales` | `ggplot2::Scale`
`coordinates` | `ggplot2::Coord`
`facet` | `ggplot2::Facet`

--

* The convention for class names is to prefix with the parent (or ancestor) `ggproto` class name and use upper camel case, e.g. `GeomPoint`.




---


class: transition 


# `Stat`

---

# Example: fit a normal distribution


🎯 Create a new layer called `stat_dist_normal()` which fits a normal density curve

.flex[
.w-50.f5[
```{r norm-dist, eval = FALSE}
# your data
df <- data.frame(x = rnorm(400, 10, 2))

# normal fit
x <- seq(min(df$x), max(df$x), 
         length.out = 1000)
fit <- data.frame(x = x,
                  y = dnorm(x, mean(df$x), sd(df$x)))

# plot
ggplot(df, aes(x)) + 
  geom_histogram(aes(y = stat(density)), bins = 30) + 
  geom_line(data = fit, color = "red", size = 2, 
            aes(x, y))
```

]
.w-50[
```{r norm-dist, fig.height = 5, echo = FALSE}
```
]
]


---

# `Stat` `ggproto`


.flex[
.w-35.pr3[
.f5[
```{r}
Stat
```
]
]
.w-65.f5[
{{content}}
]
]

--

```{r}
StatDistNormal <- ggproto("StatDistNormal", Stat,
                          compute_group = function(data, scales) {
                            x <- seq(min(data$x), max(data$x), length.out = 1000)
                            y <- dnorm(x, mean(data$x), sd(data$x))
                            data.frame(x = x, y = y)
                          },
                          required_aes = "x")

stat_dist_normal <- function(mapping = NULL, data = NULL, geom = "line",
                    position = "identity", na.rm = FALSE, show.legend = NA, 
                    inherit.aes = TRUE, ...) {
  layer(
    stat = StatDistNormal, data = data, geom = "line", position = "identity", 
    show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
    )
}
```

---

# `stat_dist_normal`

.flex[
.w-50.f5.br[
```{r stat-dist-normal, fig.width = 5, fig.height = 5}
ggplot(df, aes(x)) + 
  geom_histogram(aes(y = stat(density)), bins = 30) + 
  stat_dist_normal(color = "red", size = 2) 
```
]
.w-50.f5.pl3[
{{content}}
]
]

--

Works for facetting as well!

```{r stat-dist-normal-group, fig.width = 7, fig.height = 5}
# make some artificial group
df$group <- sample(c(0, 1), replace = TRUE, size = nrow(df))
ggplot(df, aes(x)) + 
  geom_histogram(aes(y = stat(density)), bins = 30) + 
  stat_dist_normal(color = "red", size = 2) +
  facet_wrap(~group)
```


---

class: transition 


# `Geom`


---

# Example: tic-tac-toe

🎯 Create a new layer called `geom_tictactoe()` which draws a tic-tac-toe like board

```{r, fig.height = 4, fig.width = 4}
game <- expand.grid(col = 1:3, row = 1:3)
game$move <- sample(rep(c("Alice", "Bob"), c(4, 5)))
```

.flex.mt3[
.w-30.pr3[
```{r}
game
```

]
.w-70.f4[
```{r tic-tac-toe, fig.height = 3.5, fig.width = 3.5}
ggplot(game, aes(col, row)) + 
  geom_tile(fill = "white", color = "black") +
  geom_text(aes(label = ifelse(move=="Alice", "X", "O")), 
            size = 20) +
  theme_minimal()
```
]
]


---

# `Geom` `ggproto`


.flex[
.w-35.pr3[
.f5[
```{r}
Geom
```
]
]
.w-65.f5[
{{content}}
]
]

--

```{r}
GeomTile
```

---

# `GeomTicTacToe`

.f5[
```{r}
GeomTicTacToe <- ggproto("GeomTicTacToe", GeomTile,
                         draw_panel = function(data, panel_params, coord) {
                           coords <- coord$transform(data, panel_params)
                           width <- coords$xmax - coords$xmin
                           height <- coords$ymax - coords$ymin
                           tiles <- grid::rectGrob(coords$xmin, coords$ymax, 
                                                   width = width, 
                                                   height = height, 
                                                   default.units = "native", 
                                                   just = c("left", "top"), 
                                                   gp = grid::gpar(col = coords$colour, 
                                                                   fill = alpha(coords$fill, coords$alpha), 
                                                                   lwd = coords$size * .pt, lty = coords$linetype, 
                                                                   linejoin = "mitre", lineend = "square"))
                           if(length(unique(coords$label)) != 2) {
                             stop("There should be only two players in tic-tac-toe")
                           }
                           fontsize <- min(c(height, width))
                           fontsize <- grid::convertUnit(grid::unit(fontsize, "snpc"), "pt")
                           moves <- grid::textGrob(label = factor(coords$label, labels = c("X", "O")),
                                                   x = coords$x, y = coords$y, 
                                                   gp = grid::gpar(fontsize = fontsize))
                           ggplot2:::ggname("geom_tictactoe", grid::gTree(children = grid::gList(tiles, moves)))
                         },
                         required_aes = c("x", "y", "label"))

```

]

---

# `geom_tictactoe`


```{r}
geom_tictactoe <- function(mapping = NULL, data = NULL, stat = "identity", position = "identity", 
    ..., na.rm = FALSE, show.legend = NA, inherit.aes = TRUE, color = "black", fill = "white")  {
      list( #<<
        layer(data = data, mapping = mapping, stat = stat, geom = GeomTicTacToe, 
              position = position, show.legend = show.legend, inherit.aes = inherit.aes, 
              params = list(na.rm = na.rm, color = color, fill = fill, ...)),
        theme_minimal()) #<<
    }

```


--

.flex[
.w-50[

```{r game-tictactoe, eval = FALSE}
ggplot(game, aes(col, row, 
                 label = move)) +
  geom_tictactoe() 
```
]
.w-50[
```{r game-tictactoe, fig.height = 4, fig.width = 4, echo = FALSE}
```
]
]


---

# More than the tic-tac-toe game

```{r}
checklist <- expand.grid(person = c("Alex", "Sandy"), 
                         task = c("Shop", "Work", "Exercise", "Cook"))
checklist$done <- c(TRUE, FALSE, TRUE, FALSE, TRUE, TRUE, TRUE, FALSE)
```

.flex.mt3[
.w-50.pr3[
```{r}
checklist
```

]
.w-50[
```{r gg-checklist, fig.height = 4}
ggplot(checklist, 
       aes(task, person, label = done)) +
  geom_tictactoe()
```
]
]










---

class: transition

# Resources

<br>

.f2[For more see the ["Extending ggplot2" vignette](https://cran.r-project.org/web/packages/ggplot2/vignettes/extending-ggplot2.html)]


.f2[Check out also the 3rd edition of the ggplot2 book<br>https://ggplot2-book.org/extensions.html]

.f2[And more also in documentation at .monash-white[`?ggplot2::Layout`]]



---


class: exercise middle hide-slide-number


<i class="fas fa-users"></i>

# <i class="fas fa-code"></i> Open `day1-exercise-03.Rmd`

<center>
`r countdown::countdown(15, class = "clock")`
</center>

---

```{r endslide, child="assets/endslide.Rmd"}
```
