---
title: "Advanced data visualization with R <br> Workshop Day 2"
subtitle: "Lesson 2: Digging deeper into reactive elements in shiny"
author: "Di Cook"
email: "dicook@monash.edu"
date: "8-9th Dec 2021 @ Statistical Society of Australia | Zoom"
length: "360 minutes"
color_theme: "yellow"
pdflink: "day2session2.pdf"
titleslide_bgimg: "images/d2bg3.jpg"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/animate.css"
      - "assets/styles2.css"
      - "assets/fira-code.css"
      - "assets/monash-brand.css"
      - "assets/boxes.css"
      - "assets/custom.css"
      - "assets/tachyons-addon.css"
      - "assets/scroll.css"
    self_contained: false 
    seal: false 
    chakra: 'libs/remark-latest.min.js'
    lib_dir: libs
    mathjax: "assets/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: github
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
current_file <- knitr::current_input()
```
```{r titleslide, child="assets/titleslide2.Rmd"}
```
```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "images/day2-session1/",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "100%",
  fig.retina = 3,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
library(tidyverse)
library(palmerpenguins)
library(plotly)
library(gganimate)
library(DT)
library(leaflet)
library(crosstalk)
filter <- dplyr::filter
select <- dplyr::select
```

---
class: transition middle
# Outline

Animation<br>
Linking between plots<br>
Using a `key` variable to link between plots

---
class: transition middle

# Animation


---

.pull-left[
.monash-orange2[Animation] allows for controlled changes in a plot. It can provide more information to be displayed than is possible within a single plot. It can also .monash-orange2[draw attention] to specific parts of the plot. 

Human perception is fleeting, though, and prone to .monash-orange2[forgetting] what was just displayed.  Keeping some elements persistent, possibly faint, or light coloured, can be useful for perceiving patterns of change.

Think about what part of a plot would lend itself to animation, eg a time variable, or possibly a categorical variable.
]

.pull-right[
Primary tools: `plotly`, `gganimate`

<br><br>
```{r eval=FALSE}
# Schema for animation
ggplot %>%
  animation %>%
  options (transition, 
           speed, 
           fade)
```

]
---
# ABS and Election data 

```{r absdata, echo=FALSE, eval=FALSE}
library(eechidna)
data(abs2001)
data(abs2004)
data(abs2006)
data(abs2007)
data(abs2010)
data(abs2011)
data(abs2013)
data(abs2016)
data(abs2019)
abs2001 <- abs2001 %>% 
  mutate(year = 2001) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed)
abs2004 <- abs2004 %>% 
  mutate(year = 2004) %>%
  select(year, UniqueID, Mortgage, Owned, Renting, Unemployed) %>%
  left_join(select(abs2001, UniqueID, State)) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed)
abs2006 <- abs2006 %>% 
  mutate(year = 2006) %>%
  select(year, UniqueID, Mortgage, Owned, Renting, Unemployed) %>%
  left_join(select(abs2001, UniqueID, State)) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed)
abs2007 <- abs2007 %>% 
  mutate(year = 2007) %>%
  select(year, UniqueID, Mortgage, Owned, Renting, Unemployed) %>%
  left_join(select(abs2001, UniqueID, State)) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed)
abs2010 <- abs2010 %>% 
  mutate(year = 2010) %>%
  select(year, UniqueID, Mortgage, Owned, Renting, Unemployed) %>%
  left_join(select(abs2001, UniqueID, State)) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed)
abs2011 <- abs2011 %>% 
  mutate(year = 2011) %>%
  select(year, UniqueID, Mortgage, Owned, Renting, Unemployed, Unemployed) %>%
  left_join(select(abs2001, UniqueID, State)) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed, Unemployed)
abs2013 <- abs2013 %>% 
  mutate(year = 2013) %>%
  select(year, UniqueID, Mortgage, Owned, Renting, Unemployed, Unemployed) %>%
  left_join(select(abs2001, UniqueID, State)) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed, Unemployed)
abs2016 <- abs2016 %>% 
  mutate(year = 2016) %>%
  select(year, UniqueID, Mortgage, Owned, Renting, Unemployed, Unemployed) %>%
  left_join(select(abs2001, UniqueID, State)) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed)
abs2019 <- abs2019 %>% 
  mutate(year = 2019) %>%
  select(year, UniqueID, Mortgage, Owned, Renting, Unemployed) %>%
  left_join(select(abs2001, UniqueID, State)) %>%
  select(year, UniqueID, State, Mortgage, Owned, Renting, Unemployed)
abs <- bind_rows(abs2001,
                 abs2004,
                 abs2006,
                 abs2007,
                 abs2010,
                 abs2011,
                 abs2013,
                 abs2016,
                 abs2019) %>%
  drop_na() # Electorate changes each year, and State var missing
save(abs, file=here::here("data/abs.rda"))
```

.font_small[
```{r echo=FALSE}
load(here::here("data/abs.rda"))
abs %>% 
  DT::datatable(width=1150, height=100) %>%
  DT::formatRound(columns=c('Mortgage', 'Owned', 'Renting', 'Unemployed'), digits=2)
```
]


<div class="tag center animated rubberBand" style="position:absolute;top:2%;left:12%;transform:rotate(-3deg);">
demo data for animation
</div>

.footnote[From the eechidna package, https://github.com/jforbes14/eechidna/]
---
# Animation with plotly


.left-code[
.font_small[
```{r abs_anim, results='hide', fig.show='hide'}
abs_p <- ggplot(abs, aes(
    x=Mortgage, 
    y=Owned, 
    colour=State)) + 
  geom_point(aes(frame=year, #<<
                 ids=UniqueID)) + #<<
  scale_colour_brewer("", 
     palette = "Dark2")
ggplotly(abs_p) #<<
```
]]

.right-plot[

```{r ref.label = "abs_anim", echo=FALSE, fig.width=6, fig.height=6, out.width="100%"}
```
]

---
# Animation with plotly


.left-code[
.font_small[
```{r abs_anim2, results='hide', fig.show='hide'}
abs_p <- ggplot(abs, aes(
    x=Mortgage, 
    y=Owned, 
    colour=State)) + 
  geom_point(aes(frame=year, 
                 ids=UniqueID)) + 
  scale_colour_brewer("", 
     palette = "Dark2")
ggplotly(abs_p) %>%
  animation_opts(transition=0) #<<
```
]]

.right-plot[

```{r ref.label = "abs_anim2", echo=FALSE, fig.width=6, fig.height=6, out.width="100%"}
```
]

---
# Animation with plotly


.left-code[
.font_small[
```{r abs_anim3, results='hide', fig.show='hide'}
abs_p <- ggplot(abs, aes(
    x=Mortgage, 
    y=Owned, 
    colour=State)) + 
  geom_point(aes(frame=year, 
                 ids=UniqueID)) + 
  scale_colour_brewer("", 
     palette = "Dark2")
ggplotly(abs_p) %>%
  animation_opts(frame=1000, #<<
                 transition=950, #<<
                 easing="cubic") #<<
```
]]

.right-plot[

```{r ref.label = "abs_anim3", echo=FALSE, fig.width=6, fig.height=6, out.width="100%"}
```
]

---
# Animation with plotly


.left-code[
.font_small[
```{r abs_anim4, results='hide', fig.show='hide'}
abs_p2 <- ggplot(abs, aes(
    x=Mortgage, 
    y=Owned, 
    colour=State)) + 
  geom_point(aes(frame=year)) + #<<
  scale_colour_brewer("", 
     palette = "Dark2")
ggplotly(abs_p2) %>%
  animation_opts(frame=1000, 
                 transition=950) 
```

]

The `ids` match up the points from one frame to another. If they aren't set the transition can look chaotic. 

]

.right-plot[

```{r ref.label = "abs_anim4", echo=FALSE, fig.width=6, fig.height=6, out.width="100%"}
```
]

---
# Animation with plotly


.left-code[
.font_small[
```{r abs_anim5, results='hide', fig.show='hide'}
abs_p <- ggplot(abs, aes(
    x=Mortgage, 
    y=Owned, 
    colour=State)) + 
  geom_point(aes(frame=year, 
                 ids=UniqueID)) + 
  scale_colour_brewer("", 
     palette = "Dark2")
ggplotly(abs_p) %>%
  animation_slider(hide=TRUE) #<<
```
]]

.right-plot[

```{r ref.label = "abs_anim5", echo=FALSE, fig.width=6, fig.height=6, out.width="100%"}
```
]

---
# Saving your animation

.left-code[
.font_small[
```{r}
library(htmlwidgets)
abs_ply <- ggplotly(abs_p, 
                    width=500, 
                    height=500) %>%
  animation_slider(hide=TRUE) 
saveWidget(abs_ply, "abs_ply.html", #<<
           selfcontained = F) #<<
```
]

<br><br>
To embed it in the html document:

`<iframe src="images/abs_ply.html" width="100%" height="500" </iframe>`

]

.right-plot[

<iframe src="images/abs_ply.html" width="100%" height="500" </iframe>
]
---
# Animation with gganimate

<br><br>
The `gganimate` package is recommended for more control of an animation. There are more choices for different transitions, easing, traces from previous position. 

<br>
Construction follows the same as for `plotly`: generate your ggplot, specify the animation, add options. 

---
class: transition middle

# Linking between plots

---
# Using crosstalk to link information between plots

Crosstalk extends htmlwidgets with a set of classes, functions, and conventions for implementing cross-widget interactions (currently, .monash-orange2[linked brushing] and .monash-orange2[filtering]).

.pull-left[
.font_small[
```{r crosstalk-example, results='hide', fig.show='hide'}
# Wrap data frame in SharedData
sd <- SharedData$new(
  quakes[sample(nrow(quakes), 
                100),]) #<<

# Use SharedData like a dataframe with Crosstalk-enabled widgets
bscols(
  leaflet(sd) %>% 
    addTiles() %>% 
    addMarkers(),
  datatable(sd, class="compact", 
            width="200"))
```
]]

.pull-right[

.font_small[
```{r ref.label="crosstalk-example", out.width="100%", echo=FALSE}
```
]
]

---
# Mapping out the data analysis: ABS

.font_small[
```{r echo=FALSE}
abs %>% 
  DT::datatable(width=1150, height=100) %>%
  DT::formatRound(columns=c('Mortgage', 'Owned', 'Renting', 'Unemployed'), digits=2)
```
]

---
# Mapping out the data analysis: ABS

<br>
<br>
Think about the questions that might be answered with the data

- Is the percentage of home ownership changing over the time period?
- What is the relationship between home ownership and having a mortgage?
- Is home ownership lower where there is higher unemployment?
- Is there a difference across states in percentage of renters?

---
# Trends over time

.left-code[
.font_small[
```{r abs-temporal, results='hide', fig.show='hide'}
abs_long <- abs %>%
  pivot_longer("Mortgage":"Unemployed", 
               names_to ="Statistic", 
               values_to = "Percentage")
ggplot(abs_long, aes(x=year, 
                     y=Percentage)) + 
  geom_line(aes(group=UniqueID), 
            alpha=0.5) +
  facet_wrap(~Statistic, 
             scales="free_y")
```
]]

.right-plot[

```{r ref.label="abs-temporal", echo=FALSE, fig.height=6, fig.width=6}
```
]
---
# Trends over time and linking

.left-code[
.font_small[
```{r abs-linking, results='hide', fig.show='hide'}
abs_key <- highlight_key(abs_long, #<<
                         ~UniqueID) #<<

abs_p2 <- ggplot(abs_key, #<< 
                 aes(x=year,
                     y=Percentage)) + 
  geom_line(aes(group=UniqueID), 
            alpha=0.5) +
  facet_wrap(~Statistic, 
             scales="free_y")

abs_gg2 <- ggplotly(abs_p2, 
                    height = 600, 
                    width = 500) 

highlight(abs_gg2) #<<
```
]

<div align="right">
`r anicon::nia("Click on a line", animate="bounce", anitype="hover", colour="#D93F00")`
</div>
]

.right-plot[
```{r ref.label="abs-linking", echo=FALSE}
```

]
---
# Linking control

`highlight_key()` is an implementation of `crosstalk::SharedData()`

`highlight()` controls the options for selection. 

---
# Trends over time and linking

.left-code[
.font_small[
```{r abs-linking2, results='hide', fig.show='hide'}
abs_key2<- highlight_key(abs_long, 
                         ~State) #<<

abs_p3 <- ggplot(abs_key2, 
                 aes(x=year, 
                     y=Percentage, 
                     text=State)) + #<<
  geom_line(aes(group=UniqueID), 
            alpha=0.5) +
  facet_wrap(~Statistic, 
             scales="free_y")

abs_gg3 <- ggplotly(abs_p3, 
                    height = 600, 
                    width = 500) 

highlight(abs_gg3) 
```

]

<div align="right">
`r anicon::nia("Click on a line", animate="bounce", anitype="hover", colour="#D93F00")`
</div>
]

.right-plot[
```{r ref.label="abs-linking2", echo=FALSE}
```

]
---
# Exploration vs explanation

<center>
.top10-color-box[These two displays were in support of exploration.]
</center>

<br><br><br><br>
We didn't know a lot about the data. 
<br><br>Based on the data description we identified several questions of interest. <br><br>The linked plots assisted in answering the question about temporal trend, and also relative to state. 


<!-- Exercise: Make scatterplot of two variables and link the two plots using point to point linking -->

---
# Special purpose linked plots

.s500[
.font_small[
```{r tsibbletalk, results='hide', fig.show='hide'}
library(tsibble)
library(tsibbletalk)
## ---- tourism-shared
tourism_shared <- tourism_monthly %>%
  as_shared_tsibble(
    spec = (State / Region) * Purpose)
## ---- plotly-key-tree
p_l <- plotly_key_tree(tourism_shared, 
                       height = 1100, 
                       width = 800)
## ---- tourism-series
p_tr <- tourism_shared %>%
  ggplot(aes(x = Month, y = Trips)) +
  geom_line(aes(group = Region), alpha = .5, size = .4) +
  facet_wrap(~ Purpose, scales = "free_y") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")
## ---- tourism-scatter
library(feasts)
tourism_feat <- tourism_shared %>%
  features(Trips, feat_stl)
p_br <- tourism_feat %>%
  ggplot(aes(x = trend_strength, y = seasonal_strength_year)) +
  geom_point(aes(group = Region), alpha = .8, size = 2)
## ---- tourism-multi
subplot(p_l,
  subplot(
    ggplotly(p_tr, tooltip = "Region", 
             width = 700, height = 800),
    ggplotly(p_br, tooltip = "Region", 
             width = 700, height = 800),
    nrows = 2),
  widths = c(.4, .6)) %>%
  highlight(dynamic = TRUE)
```
]]

<div class="tag center animated rubberBand" style="position:absolute;top:2%;left:12%;transform:rotate(-3deg);">
demo 
</div>
---

class: exercise middle hide-slide-number


<i class="fas fa-users"></i>

# <i class="fas fa-code"></i> Open `day2-exercise-02.Rmd`

<center>
`r countdown::countdown(15, class = "clock")`
</center>
---
# Learning more

- Sievert (2019) [Interactive web-based data visualization with R, plotly, and shiny](https://plotly-r.com)
- [RStudio's htmlwidgets gallery](http://gallery.htmlwidgets.org)
- [RStudio's crosstalk guide](https://rstudio.github.io/crosstalk/)
- [tsibbletalk](https://github.com/earowang/tsibbletalk)

---

```{r endslide, child="assets/endslide.Rmd"}
```

