---
title: "Day 1 Exercise 2: Harnessing `grid` on `ggplot2` outputs"
output:
  html_document:
    toc: true
    toc_float: true
---

Push the `knit` button!

If the image is not showing up, first open the corresponding solution file (in the same location as this file) and `knit`. 


```{r setup, child="exercise-setup.Rmd", eval = file.exists("exercise-setup.Rmd")}
```

```{r pkgs, message = FALSE, warning = FALSE}
library(ggplot2)
library(grid)
```


### Exercise 2.1: Line of best fit

* Change the best line of fit below to a red color using `grid` only


```{r day1-exercise-02-01A, fig.width = 3, fig.height = 3, fig.align = "center"}
ggplot(cars, aes(speed, dist)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) + 
  ggtitle("Line of best fit")

grid.force()
grid.ls()
```

* Draw a line using `grid` that points from the title to the line of best fit in the above plot 


### Exercise 2.2: Side-by-side plots with annotations

* Draw the plots below side-by-side such that the dimension of the right plot is half of the one on the left using `grid` (some of the code has been written for you )
* Then draw a red circle around the right plot

```{r day1-exercise-02-02A, fig.width = 4, fig.height = 3, fig.align = "center"}
g1 <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_point() + 
  ggtitle("Sepal")

g2 <- ggplot(iris, aes(Petal.Length, Petal.Width)) +
  geom_point() + 
  ggtitle("Petal")

g1 
g2
```

```{r day1-exercise-02-02As, eval = FALSE}
gr1 <- ggplot_gtable(ggplot_build(g1))
gr2 <- ggplot_gtable(ggplot_build(g2))
vp1 <- viewport(width = 0.5, height = 0.5, x = 0.25)
vp2 <- viewport(width = 0.25, height = 0.25, x = 0.75)
...
```



